---
title: "Flat price exploratory 2"
output: html_notebook
---

```{r}
library(tidyverse)
library(janitor)
library(glue)
library(lubridate)

theme_set(theme_light())
```

## Read in data
```{r}
# *****************************************************************************
# read in transaction data for existing flats (this excludes new build flats)
# *****************************************************************************

exist_flat_trans <- readxl::read_xls("data/HPSSA Dataset 1 - Number of residential property sales by MSOA/HPSSA Dataset 1 - Number of residential property sales by MSOA.xls", sheet = "3e", skip = 5) %>% 
  
  # make naming of variables more consistent
  clean_names() %>% 
  
  # remove erroneous column created when reading xls file
  select(-x105)

# *****************************************************************************
# read in transaction data for all property types (this includes existing flats)
# *****************************************************************************

all_housing_trans <-  readxl::read_xls("data/HPSSA Dataset 1 - Number of residential property sales by MSOA/HPSSA Dataset 1 - Number of residential property sales by MSOA.xls", sheet = "1a", skip = 5) %>% 
  
  # make naming of variables more consistent
  clean_names() %>% 
  
  # remove erroneous column created when reading xls file
  select(-x105)
```
## Process data

The BSF data is for England only. So, I am only going to want to look at house transactions for England too.

```{r}
# *****************************************************************************
# focus down on England data only
# *****************************************************************************

# for existing flats data
exist_flat_trans_eng <- exist_flat_trans %>% 
  filter(str_detect(local_authority_code, "^E")) 

# for all properties data
all_housing_trans_eng <- all_housing_trans %>% 
  filter(str_detect(local_authority_code, "^E")) 
```


## Tidy data

```{r}
# *****************************************************************************
# tidy the data ahead of analysis
# *****************************************************************************

#' A function to tidy data read in from an individual sheet within the .xls spreadsheet:
#' "HPSSA Dataset 1 - Number of residential property sales by MSOA.xls"
#'
#' @param df: a dataframe
#' @param df: a string to identify the type of transactions counted 
#'
#' @return: a tidied tibble of the data

tidy_housing_trans_data <- function(df, trans_col_name){
  
  # tidy data
  df %>% 
    
    # deal with column names that are actually values of a variable (year_ending)
    pivot_longer(cols = year_ending_dec_1995:year_ending_sep_2020, 
                 names_to = "year_ending", 
                 values_to = trans_col_name) %>% 
    
    # process year_ending strings to create a date column
    mutate(date = str_replace(year_ending, "year_ending_", "")) %>% 
    separate(date, into = c("month", "year"), sep = "_") %>% 
    mutate(date = lubridate::my(glue("{month} {year}"))) %>% 
    
    # remove variables that are no longer needed
    select(-month, -year, -year_ending)
  
}

# tidy both the flat and all housing datasets
exist_flat_tidy <- tidy_housing_trans_data(exist_flat_trans_eng, "num_trans_ex_flats")
all_housing_tidy <- tidy_housing_trans_data(all_housing_trans_eng, "num_trans_all_housing")

# *****************************************************************************
# join the datasets for different transactions types analysis
# *****************************************************************************
transactions <- all_housing_tidy %>% 
  left_join(exist_flat_tidy) %>% 
  relocate(date, .before = msoa_code)
```

## Focus down on data of interest

Data reported for the year ending June 2017 is the last set of observations where it is improbable that the Grenfell Tower Fire could have had major impacts on transaction numbers.

```{r}
trans_focus <- transactions %>% 
  # remove data from prior to the Grenfell Fire
  filter(date >= lubridate::my("Jun 2016")) %>% 
  
  # create a new variable to allow comparison of existing flats with the rest of
  # the market
  mutate(num_trans_not_ex_flats = num_trans_all_housing - num_trans_ex_flats)
```

## Create the additional variables needed for plotting and analysis
```{r}
trans_init_plot <- trans_focus %>% 
  
  # the data needs tidying again after joining
  pivot_longer(cols = starts_with("num_trans"),
               names_to = "trans_type",
               values_to = "num_trans") %>% 
  
  # remove unnecessary text within newly created column
  mutate(trans_type = str_replace(trans_type, "num_trans_", "")) %>% 
  
  # calculate transaction counts for England as a whole
  group_by(date, trans_type) %>% 
  summarise(num_trans = sum(num_trans))
  
  
  
```

## Initial plotting

```{r}
ggplot(trans_init_plot, aes(date, num_trans,
                            colour = trans_type)) +
  geom_point() +
  geom_line() +

  expand_limits(y = 0) +
  
  labs(title = "Existing flats make up a small proportion of over housing transactions",
       subtitle = "So it is difficult to compare trends in transaction numbers ...")
  
```
So, I look at percentage change in transactions numbers after Grenfell. Using the year ending June 2017 transaction numbers as a baseline.

```{r}
# ****************************************************************
# calculate the baseline against which percent change can be measured
# ****************************************************************
baseline_date <- lubridate::date("2016-06-01")

baseline_trans_nums <- trans_init_plot %>% 
  filter(date == baseline_date)

bl_all_housing <- baseline_trans_nums[[1,3]]
bl_ex_flats <- baseline_trans_nums[[2,3]]
bl_not_ex_flats <- baseline_trans_nums[[3,3]]

# prepare percent change data for plotting
pc_plot_df <-  trans_init_plot %>%
  
  # data needs to be wide for percent change calculation
  pivot_wider(names_from = trans_type, values_from = num_trans) %>% 
  
  # calculate percentage change
  mutate(pc_all_housing = ((all_housing - bl_all_housing) / bl_all_housing) * 100,
         pc_ex_flats = ((ex_flats - bl_ex_flats) / bl_ex_flats) * 100,
         pc_not_ex_flats = ((not_ex_flats - bl_not_ex_flats) / bl_not_ex_flats) * 100) %>% 
  
  # (re-)tidy data for plotting
  pivot_longer(cols = starts_with("pc"), 
               names_to = "trans_type", 
               values_to = "percent_change") %>%
  
  # remove unnecessary text from entries in trans_type column
  mutate(trans_type = str_replace(trans_type, "pc_", "")) %>% 
  
  # remove unnecessary columns
  select(-all_housing, -ex_flats, -not_ex_flats)
```

```{r}
ggplot(pc_plot_df, aes(date, percent_change,
                       colour = trans_type)) +
  geom_vline(xintercept = lubridate::date("2017-06-14")) +
  geom_point() +
  geom_line() +
  annotate("text", x = lubridate::date("2017-06-14"), y = -40,
           label = "Grenfell tower fire", hjust = -0.05) +
  labs(title = "By Sept 2020 the number of exisiting flats per year has reduced over 45%\nsince June 2016",
       subtitle = "The number of existing flat sales per year was already falling before the\nGrenfell Tower fire")
```

## Looking a percent change in number of flat sales at the local authority scale

```{r}
la_trans <- trans_focus %>% 
  
  # calculate local authority totals for transaction count type
  group_by(date, local_authority_name, local_authority_code) %>% 
  summarise(num_trans_all_housing = sum(num_trans_all_housing),
            num_trans_ex_flats = sum(num_trans_ex_flats),
            num_trans_not_ex_flats = sum(num_trans_not_ex_flats)) %>%
  ungroup() %>% 
  
    # the data needs tidying again after calculating local authority totals
  pivot_longer(cols = starts_with("num_trans"),
               names_to = "trans_type",
               values_to = "num_trans") %>% 
  
  # remove unnecessary text within newly created column
  mutate(trans_type = str_replace(trans_type, "num_trans_", "")) 
  

baseline_date <- lubridate::date("2017-06-01")

la_baseline_trans_nums <- la_trans %>% 
  filter(date == baseline_date) %>% 
  rename(baseline_num_trans = num_trans) %>% 
  select(-date)

la_pc <- la_trans %>% 
  left_join(la_baseline_trans_nums) %>% 
  mutate(pc = ((num_trans - baseline_num_trans)/ baseline_num_trans),
         change_in_trans = num_trans - baseline_num_trans) %>% 
  
  filter(date == date("2020-06-01") & trans_type == "ex_flats" & baseline_num_trans > 100) %>% 
  arrange(change_in_trans)

View(la_pc)





```

```{r}

```

